Copulas and their potential for ecology

### 2019-10-01 Paper.Rmd
more comments refining what I will next do to Q2 sections

To explore the importance of nonlinear dynamics for our proposed mechanism, 
we also considered a
nonlinear population model $P_i(t+1)=P_i(t)\exp \left(r\left(1-P_i(t)/K  \right)+\sigma\epsilon_i(t)\right)$, using $r=0.5$, $K=100$, and $\sigma=0.1$ or 
$\sigma=1$. When $\sigma=0.1$, population dynamics stay close to the carrying capacity, 
$K$, and the nonlinearities of the model have limited influence on dynamics. When
$\sigma=1$, model dynamics are strongly nonlinear. We refer to these as
the weak-noise and strong-noise cases, though the importance of the noise here
is that, when it is strong, it brings the nonlinearities of the model into play. 
<!--***DAN: Didn't we use another value of r as well, and present results
on it? If so, probably should mention it here, in part so the exploration
seems a bit broader?-->

For each of the model setups above, for each of several copula families, 
<!--***DAN: I currently use C, SC, G, SG, J, SJ. Do I really need all those? 
Surely it would be enough to just do C and SC and then state that the others 
give similar results? And if I move one of the result figure from this section
into the main text, that will mean only one such fig in the sup mat, at least
for AR(1)? I mean, right now we have 18 figure in sup mat from this analysis,
and only part of one of them in main text! See if the results are actually similar across
S/SC, G/SG, J/SJ, and if so consider keeping only the S/SC results in either the
main text or the sup mat, and just stating the others were similar. Then you would presumably
move the full C fig into the main text, replacing the current partial version.-->
for each $\tau=0.1,0.2,\ldots,0.9$, and for each of
$50$ replicate simulations, we generated 5000 noise pairs 
$(\epsilon_1(t),\epsilon_2(t))$ from the
bivariate random variable with the given copula family and the given Kendall
correlation $\tau$.
We then used this noise to drive the model, and retained both the
noise and population values for the final 500 
time steps. For each simulation, the following
statistics were then computed for both noises and populations: 
Pearson, Spearman and Kendall correlations, $\cor_l$, $\cor_u$, $\Ps_l$, $\Ps_u$, 
$\Dsq_l$, $\Dsq_u$, $\cor_l - \cor_u$, $\Ps_l - \Ps_u$, and $\Dsq_u - \Dsq_l$. 
Values were plotted against $\tau$ for noises and populations. If the 
hypothesis from the Introduction was correct that characteristics of 
the copula structure of spatial dependence in an
ecological variable can be inherited from the copula structure of spatial 
dependence in an environmental variables, then 
plots should be similar for populations and noises.

@@

Our model with asymmetric environmental effects 
produces outputs with visually apparent asymmetry 
of tail dependence, to an extent that depended on the value of $b$;
thus this is a mechanism that can produce non-normal copula structure and
asymmetric tail dependence across space in ecological variables.
<!--***DAN: move the following to the end of this part-->
This result is represented in Fig. \ref{MasterFigure} as the solid box around 
"Nonlinear environmental effects, Liebig's law" and the arrows labelled "A".
It is explained in the Discussion why the box and some of those arrows 
are solid, instead of dashed, although the results we 
present in this section are theoretical.
<!--***DAN: move the above to the end of this part-->
For $b=0.1$ (Fig. \ref{fig_ms_asym_b_0.1}) and $b=-0.1$ 
(Fig. \ref{SM-fig_ms_asym_b_-0.1}), 
for both $g_1$ and $g_2$,
asymmetry of tail dependence was strong; for 
$b=\pm 0.5$, asymmetry was weaker but still apparent
(Figs \ref{SM-fig_ms_asym_b_0.5} - 
\ref{SM-fig_ms_asym_b_-0.5}).
<!--***DAN: combin all 3 sup mat figs on this subject into one sup mat fig-->
Lower-tail (respectively, upper-tail) spatial dependence in 
effects of noise ($g_1$, respectively, $g_2$) 
produced lower-tail (respectively, upper-tail) spatial 
dependence in the ecological variable.
Model selection results (Table \ref{SM-tab_summary_stat_asym_sens}) confirmed that
ecological outputs had non-normal copula structure, but did not always reflect 
asymmetry of tail dependence, perhaps because those methods measured
dependence in the extreme tails. Results using our nonparametric statistics, however,
reflected the visually apparent asymmetry (Tables \ref{SM-tab_stat_npa_asym}). 
<!--***DAN: Simplify as follows, maybe:
1) get rid of table S10 (you will have to change the methods so it no longer says we
do out model selection analysis here)
2) simplify table S11 so it only shows the cor results (also means a change to methods)
and then maybe bring the resulting table to main text?
-->

>>>>

To explore the importance of nonlinear dynamics for our proposed mechanism, 
we also considered a
nonlinear population model $P_i(t+1)=P_i(t)\exp \left(r\left(1-P_i(t)/K  \right)+\sigma\epsilon_i(t)\right)$, using $r=0.5$, $K=100$, and $\sigma=0.1$ or 
$\sigma=1$. When $\sigma=0.1$, population dynamics stay close to the carrying capacity, 
$K$, and the nonlinearities of the model have limited influence on dynamics. When
$\sigma=1$, model dynamics are strongly nonlinear. We refer to these as
the weak-noise and strong-noise cases, though the importance of the noise here
is that, when it is strong, it brings the nonlinearities of the model into play. 
<!--***DAN: Didn't we use another value of r as well, and present results
on it? If so, probably should mention it here, in part so the exploration
seems a bit broader? Well, we mentioned that we did those and we verbally summarize
results but we did not show any. So maybe leave things as is in this respect.-->

For each of the model setups above, for each of several copula families, 
<!--***DAN: I currently use C, SC, G, SG, J, SJ. Do I really need all those? 
Surely it would be enough to just do C and SC and then state that the others 
give similar results? And if I move one of the result figure from this section
into the main text, that will mean only one such fig in the sup mat, at least
for AR(1)? I mean, right now we have 18 figure in sup mat from this analysis,
and only part of one of them in main text! See if the results are actually similar across
S/SC, G/SG, J/SJ, and if so consider keeping only the S/SC results in either the
main text or the sup mat, and just stating the others were similar. Then you would presumably
move the full C fig into the main text, replacing the current partial version.
So instead of having 18 figs in sup mat and half a fig in main text, you would
have 5 figs in sup mat and 1 in main text, and then you would just say "we did the
same thing for G, SG, J, SJ and got similar results."
So let's check, are results, broadly speaking, independent of which of the 6 copulas
I used?:
A) for AR(1) models, yes
B) for Ricker, weak noise, yes
C) for Ricker, strong noise, its more complicated. It's generally true that you get
more approximate agreement between noise and pop stats, with some rather glaring disagreement.
But which stats disagree the most depends on which cop we used!
So I guess we can still just display the C and SC results, as planned above, and explain
in text.
-->
for each $\tau=0.1,0.2,\ldots,0.9$, and for each of
$50$ replicate simulations, we generated 5000 noise pairs 
$(\epsilon_1(t),\epsilon_2(t))$ from the
bivariate random variable with the given copula family and the given Kendall
correlation $\tau$.
We then used this noise to drive the model, and retained both the
noise and population values for the final 500 
time steps. For each simulation, the following
statistics were then computed for both noises and populations: 
Pearson, Spearman and Kendall correlations, $\cor_l$, $\cor_u$, $\Ps_l$, $\Ps_u$, 
$\Dsq_l$, $\Dsq_u$, $\cor_l - \cor_u$, $\Ps_l - \Ps_u$, and $\Dsq_u - \Dsq_l$. 
Values were plotted against $\tau$ for noises and populations. If the 
hypothesis from the Introduction was correct that characteristics of 
the copula structure of spatial dependence in an
ecological variable can be inherited from the copula structure of spatial 
dependence in an environmental variables, then 
plots should be similar for populations and noises.

@@

Our model with asymmetric environmental effects 
produces outputs with visually apparent asymmetry 
of tail dependence, to an extent that depended on the value of $b$;
thus this is a mechanism that can produce non-normal copula structure and
asymmetric tail dependence across space in ecological variables.
<!--***DAN: move the following to the end of this part-->
This result is represented in Fig. \ref{MasterFigure} as the solid box around 
"Nonlinear environmental effects, Liebig's law" and the arrows labelled "A".
It is explained in the Discussion why the box and some of those arrows 
are solid, instead of dashed, although the results we 
present in this section are theoretical.
<!--***DAN: move the above to the end of this part-->
For $b=0.1$ (Fig. \ref{fig_ms_asym_b_0.1}) and $b=-0.1$ 
(Fig. \ref{SM-fig_ms_asym_b_-0.1}), 
for both $g_1$ and $g_2$,
asymmetry of tail dependence was strong; for 
$b=\pm 0.5$, asymmetry was weaker but still apparent
(Figs \ref{SM-fig_ms_asym_b_0.5} - 
\ref{SM-fig_ms_asym_b_-0.5}).
<!--***DAN: Shyamolina, can you please combine Figs S24-S26 into one single figure
with 6 panels? Please also search for citations in the main text and sup mat that 
used to refer to the three figures and replace them with a citation to the
new, combined figure.-->
Lower-tail (respectively, upper-tail) spatial dependence in 
effects of noise ($g_1$, respectively, $g_2$) 
produced lower-tail (respectively, upper-tail) spatial 
dependence in the ecological variable.
Model selection results (Table \ref{SM-tab_summary_stat_asym_sens}) confirmed that
ecological outputs had non-normal copula structure, but did not always reflect 
asymmetry of tail dependence, perhaps because those methods measured
dependence in the extreme tails. Results using our nonparametric statistics, however,
reflected the visually apparent asymmetry (Tables \ref{SM-tab_stat_npa_asym}). 
<!--***DAN: Simplify as follows, maybe:
1) get rid of table S10 (you will have to change the methods so it no longer says we
do our model selection analysis here)
2) simplify table S11 so it only shows the cor results (also means a change to methods)
and then maybe bring the resulting table to main text?
-->

<!--***DAN: Shyamolina, can you please simplify Table S11 so it only shows the cor
results? That also means the third column will be unnecessary. Please do this as follows:
1) keep but comment out the code that produces the old table; 2) write new code (presumably
starting from a copy of the old code) that makes the new, simpler version of the table. 
That will make it very easy to produce the old version again if we ever want to.-->
